---
title: "exposure-duration-histogram"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{exposure-duration-histogram}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(metalite.sl)
library(dplyr)
```

# Create Exposure Duration Histogram

The plotly_exp_duration() function provides an interactive visualization of subject-level exposure duration in clinical trials. This histogram helps assess how long participants remained on treatment, with configurable duration categories such as "≥1 day", "≥7 days", "≥28 days", and so on.

Using pre-processed output from prepare_exp_duration() (and optionally extend_exp_duration()), plotly_exp_duration() generates an intuitive, browser-based plot built with Plotly. Users can customize the display type (e.g., counts or proportions), color, tooltip summary statistics, and axis labels.

-  `meta_sl_exposure_example`: create example exposure metadata (`meta` object) for demonstration or testing purposes.

-  `prepare_exp_duration`: process subject-level exposure data to calculate treatment duration and prepare it for further analysis.

-  `extend_exp_duration`: categorize duration into user-defined time windows (e.g., ≥7 days, ≥12 weeks) and add group-level summaries.

-  `plotly_exp_duration`: generate an interactive histogram plot to visualize exposure duration distribution across subjects or treatment arms.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/exposure0histogram.pdf")
```

## 1. Load up a metadata
## meta_sl_exposure_example()
There are two steps in `meta_sl_exposure_example` function in order to build the metadata (`meta` object): processing the ADaM dataset and save meta information for A&R reporting.

Step1: Load up existing ADEXSUM, the ADaM dataset for Drug Exposrue Summary Data, that contains:

- Duration summed up by STUDYID SITENUM USUBJID SUBJID APERIOD EXTRT ADOSEFRM PARAMCD.
- The exposure data subset by `upcase(trim(left(paramcd))) = "TRTDURD"`.
- The exposure duration `adexsum$AVAL` for all participants.
- Duration category assigned with `adexsum$EXDURGR` i.e.">=1 day", ">=7 days",">=28 days", ">=12 weeks" and ">=24 weeks".
```{r}
library(metalite.ae)
data("metalite_ae_adexsum", package = "metalite.ae") 
adexsum <- metalite_ae_adexsum
```
Step2: Save analysis plan and metadata(parameter and analysis) information, then build meta object.

```{r}
plan <- metalite::plan(
  analysis = "exp_dur", population = "apat",
  observation = "apat", parameter = "expdur"
)

meta <- metalite::meta_adam(
  population = adexsum,
  observation = adexsum
) |>
  metalite::define_plan(plan) |>
  metalite::define_population(
    name = "apat",
    group = "TRTA",
    subset = quote(APERIOD == 1 & AVAL > 0)
  ) |>
  metalite::define_parameter(
    name = "expdur",
    subset = PARAMCD == "TRTDURD",
    var = "AVAL",
    label = "Exposure Duration (Days)",
    vargroup = "EXDURGR"
  ) |>
  metalite::define_analysis(
    name = "exp_dur",
    title = "Summary of Exposure Duration",
    label = "exposure duration table"
  ) |>
  metalite::meta_build()
```


## 2. Analysis preparation
## prepare_exp_duration()

The input of the function `prepare_exp_duration()` is a `meta` object created by the metalite package. The resulting output comprises a collection of raw datasets for analysis and reporting.

This function takes raw data (e.g., participant-level data with exposure durations and treatment groups).
It calculates summary statistics (counts, proportions, medians, etc.) grouped by treatment group.
It also creates exposure duration categories (e.g., >=1 day, >=7 days, etc.) by binning or categorizing the raw exposure duration variable.
The output is a structured dataset (outdata) that contains:
Counts and proportions of participants in each treatment group.
Summary statistics for the exposure duration variable within each treatment group.
However, at this stage, the statistics are grouped only by treatment group, not yet broken down by exposure duration categories.

```{r}
outdata <- prepare_exp_duration(meta)
outdata
```

Number of participants in population

```{r}
outdata$n[, 1:5]
```

Number of participants in each duration category

```{r}
charn <- data.frame(outdata$char_n[1])
head(charn[, 1:5], 6)
```

Proportion of participants in each duration category

```{r}
charp <- data.frame(outdata$char_prop[1])
head(charp[, 1:5], 6)
```

Statistical summary of exposure duration for each treatment

```{r}
chars <- data.frame(outdata$char_n[1])
tail(chars[, 1:5], 8)
```
## extend_exp_duration()
This function takes the output of prepare_exp_duration() and further processes the data to create exposure duration categories explicitly.
It applies the specified duration category cutoffs (e.g., >=1 day, >=7 days, etc.) to the exposure duration variable.
It calculates counts, proportions, and summary statistics for each exposure duration category within each treatment group.
This means now we have a two-dimensional grouping:
By treatment group (e.g., Placebo, Low Dose and High Dose).
By exposure duration category (e.g., >=1 day, >=7 days).
The output outdata now contains tables like:
char_n: counts by exposure duration category and treatment group.
char_prop: proportions by exposure duration category and treatment group.
char_stat_groups: summary statistics by exposure duration category and treatment group.


```{r}
outdata <- meta |>
  prepare_exp_duration() |>
  extend_exp_duration(
    duration_category_list = list(c(1, NA), c(7, NA), c(28, NA), c(12 * 7, NA), c(24 * 7, NA)),
    duration_category_labels = c(">=1 day", ">=7 days", ">=28 days", ">=12 weeks", ">=24 weeks")
  )
outdata
```

```{r}
all_stats <- bind_rows(outdata$char_stat_groups, .id = "duration_category")

all_stats
```

## plotly_exp_duration()
The function takes the fully prepared outdata from the above steps to simply visualizes these pre-calculated statistics in an interactive way.
It reshapes the counts and proportions tables from wide to long format, so each row corresponds to a specific combination of:
Exposure duration category (e.g., >=7 days).
Treatment group (e.g., Placebo).
It uses these reshaped tables to create stacked bar charts or grouped bar charts that show:
The number or proportion of participants in each exposure duration category for each treatment group.
It also uses the summary statistics (char_stat_groups) to create hover text that shows detailed statistics for each bar (i.e., for each treatment group × exposure duration category).
The interactive dropdown lets you switch between different views (e.g., cumulative exposure duration, exclusive categories, horizontal bars).

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
outdata |>
  plotly_exp_duration(
    color = NULL,
    display = "n",
    display_total = TRUE,
    plot_group_label = "Treatment Group",
    plot_category_label = "Exposure Duration",
    hover_summary_var = c("mean", "min", "max"),
    width = 800,
    height = 400
  )
```

