---
title: "Build an Interactive Baseline Characteristic Table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build an Interactive Baseline Characteristic Table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message=FALSE}
library(r2rtf)
library(metalite)
library(metalite.ae)
library(metalite.sl)
```

There are 2 key metadata types:

- metadata for the baseline characteristic table
- metadata for the AE subgroup specific table

# Build metadata {.tabset}

## Metadata for baseline characteristic table

The code below is the same as `meta_sl_example()`.

```{r}
adsl <- r2rtf::r2rtf_adsl
adsl$TRTA <- adsl$TRT01A
adsl$TRTA <- factor(adsl$TRTA,
  levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"),
  labels = c("Placebo", "Low Dose", "High Dose")
)

meta <- meta_adam(
  population = adsl,
  observation = adsl
) |>
  define_plan(plan = plan(
    analysis = "base_char", population = "apat",
    observation = "apat", parameter = "age;gender;race"
  )) |>
  define_population(
    name = "apat",
    group = "TRTA",
    subset = quote(SAFFL == "Y"),
    var = c("USUBJID", "TRTA", "SAFFL", "AGEGR1", "SEX", "RACE")
  ) |>
  define_observation(
    name = "apat",
    group = "TRTA",
    subset = quote(SAFFL == "Y"),
    var = c("USUBJID", "TRTA", "SAFFL", "AGEGR1", "SEX", "RACE")
  ) |>
  define_parameter(
    name = "age",
    var = "AGE",
    label = "Age (years)",
    vargroup = "AGEGR1"
  ) |>
  define_parameter(
    name = "gender",
    var = "SEX",
    label = "Gender"
  ) |>
  define_parameter(
    name = "race",
    var = "RACE",
    label = "Race"
  ) |>
  define_analysis(
    name = "base_char",
    title = "Participant Baseline Characteristics by Treatment Group",
    label = "baseline characteristic table"
  ) |>
  meta_build()
```

## A metadata of the AE subgroup specific analysis

In this vignette, we will directly use the metadata built by `meta_ae_example()`.

```{r}
meta_ae <- meta_ae_example()
```

# Build a reactable {.tabset}

## Baseline characteristic table + Participants With Drug-Related AE

```{r, eval = TRUE}
react_base_char(
  metadata_sl = meta,
  metadata_ae = meta_ae,
  ae_subgroup = c("age", "race", "gender"),
  ae_specific = "rel",
  width = 1200
)
```

## Baseline characteristic table + Participants With Serious AE

```{r, eval=TRUE}
react_base_char(
  metadata_sl = meta,
  metadata_ae = meta_ae,
  ae_subgroup = c("age", "race", "gender"),
  ae_specific = "ser",
  width = 1200
)
```

## Customization

If you want to capitalize the first letter of each word in the "RACE" (Black or African American) or any other character variable, 
you can customize the `react_base_char` function at the beginning of the code.

```{r, eval = FALSE}
# function to capitalize the first letter of each word (proper case)
capitalize_words <- function(x) {
  # Define a vector of words to be lower-cased for minor words like articles, conjunctions, and propositions
  lowercase_exceptions <- c(
    "a", "an", "the", "and", "or", "but", "for", "nor", "on", "so",
    "yet", "at", "to", "by", "with", "in", "of", "as", "up", "with"
  )

  sapply(x, function(sentence) {
    # Split the sentence into words
    words <- strsplit(sentence, " ")[[1]]

    # Capitalize the first letter of each word, except for exceptions
    capitalized_words <- sapply(seq_along(words), function(i) {
      word <- words[i]
      if (i == 1 || !(tolower(word) %in% lowercase_exceptions)) {
        # Capitalize the first letter if it's the first word or not an exception
        paste0(toupper(substr(word, 1, 1)), tolower(substr(word, 2, nchar(word))))
      } else {
        # Return the word in lowercase if it's an exception
        tolower(word)
      }
    })

    # Combine the capitalized words back into a single string
    paste(capitalized_words, collapse = " ")
  })
}
```
If you want to capitalize only the first letter of "RACE" (e.g., Black or african american) or any other character variable, 
you can customize the `react_base_char` function at the beginning of the code.


```{r, eval = FALSE}
# function to capitalize the first letter of a string that has multiple words
capitalize_words <- function(x) {
  sapply(x, function(word) {
    paste0(toupper(substr(word, 1, 1)), tolower(substr(word, 2, nchar(word))))
  })
}
```
The first function capitalizes the first letter of each word, while the second function capitalizes only the first letter of a string that contains multiple words.

Use either of the “capitalize_words” functions above to meet your needs.  Example below is the implementation of the first function.

```{r, eval = FALSE}
# 1) In "data_population": extract the RACE values as a character vector
race_values_pop <- metadata_sl[["data_population"]]$RACE # Use $ to get a vector

# Capitalize the race values
metadata_sl[["data_population"]]$RACE <- capitalize_words(race_values_pop) # Assign back as a vector

# 2) In "data_observation": extract the RACE values as a character vector
race_values_obs <- metadata_sl[["data_observation"]]$RACE # Use $ to get a vector

# Capitalize the race values
metadata_sl[["data_observation"]]$RACE <- capitalize_words(race_values_obs) # Assign back as a vector
```
