---
title: "Getting Started with metalite.sl"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Getting Started with metalite.sl}
---

```{r}
library(metalite)
library(metalite.sl)
```

# Build a metadata
```{r}
adsl <- r2rtf::r2rtf_adsl
adsl$TRTA <- adsl$TRT01A
adsl$TRTA <- factor(adsl$TRTA,
  levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"),
  labels = c("Placebo", "Low Dose", "High Dose")
)

# Create a variable EOSSTT indicating the end of end of study status
adsl$EOSSTT <- sample(x = c("Participants Ongoing", "Discontinued"),
                      size = length(adsl$USUBJID), 
                      prob = c(0.8, 0.2), replace = TRUE)
# Create a variable EOTSTT1 indicating the end of treatment status part 1
adsl$EOTSTT1 <- sample(x = c("Completed", "Discontinued"),
                      size = length(adsl$USUBJID), 
                      prob = c(0.85, 0.15), replace = TRUE)
```

```{r}
plan <- plan(
    analysis = "base_char", population = "apat",
    observation = "apat", parameter = "age;gender;race"
  ) |>
    add_plan(
      analysis = "base_char_subgroup", population = "apat",
      observation = "apat", parameter = "age"
  ) |>
  add_plan(
    analysis = "disp", population = "apat",
    observation = "apat", parameter = "disposition;medical-disposition"    
  ) |>
  add_plan(
    analysis = "trt_compliance", population = "apat",
    observation = "apat", parameter = "comp8;comp16;comp24"    
  ) 
```

```{r}
meta <- meta_adam(
  population = adsl,
  observation = adsl
) |>
  define_plan(plan = plan) |>
  define_population(
    name = "apat",
    group = "TRTA",
    subset = quote(SAFFL == "Y"),
    var = c("USUBJID", "TRTA", "SAFFL", "AGEGR1", "SEX", "RACE", "EOSSTT", "EOTSTT1", "COMP8FL", "COMP16FL", "COMP24FL")
    
  ) |>
  define_parameter(
    name = "age",
    var = "AGE",
    label = "Age (years)",
    vargroup = "AGEGR1"
  ) |>
  define_parameter(
    name = "gender",
    var = "SEX",
    label = "Gender"
  ) |>
  define_parameter(
    name = "race",
    var = "RACE",
    label = "Race"
  ) |>
  define_parameter(
    name = "disposition",
    var = "EOSSTT",
    label = "Trial Disposition"
  ) |>
  define_parameter(
    name = "medical-disposition",
    var = "EOTSTT1",
    label = "Participant Study Medication Disposition"
  ) |>
  metalite::define_parameter(
    name = "comp8",
    var = "COMP8FL",
    label = "Compliance (Week 8)",
  ) |>
  metalite::define_parameter(
    name = "comp16",
    var = "COMP16FL",
    label = "Compliance (Week 16)",
  ) |>
  metalite::define_parameter(
    name = "comp24",
    var = "COMP24FL",
    label = "Compliance (Week 24)",
  ) |>
  define_analysis(
    name = "base_char",
    title = "Participant Baseline Characteristics by Treatment Group",
    label = "baseline characteristic table"
  ) |>
  define_analysis(
    name = "base_char_subgroup",
    title = "Participant by Age Category and Sex",
    label = "baseline characteristic sub group table"
  ) |>
  define_analysis(
    name = "trt_compliance",
    title = "Summary of Treatment Compliance",
    label = "treatment compliance table"
  ) |>
  define_analysis(
    name = "disp",
    title = "Disposition of Participant",
    label = "disposition table"
  ) |>
  meta_build()
```

# Create Baseline Characteristic Table
## Analysis preparation

```{r}
outdata <- meta |>
  prepare_base_char(
    parameter = "age;gender;race"
  )

outdata

outdata$n
outdata$char_n
outdata$char_var
outdata$char_prop
```

## Format output

```{r}
outdata <- outdata |> format_base_char(display_col = c("n", "prop", "total"), digits_prop = 2)

outdata$tbl
```

```{r}
outdata <- outdata |> format_base_char(display_col = c("n", "prop"), digits_prop = 3)

outdata$tbl
```

## Output as RTF

```{r}
outdata |> rtf_base_char(
  source = "Source: [CDISCpilot: adam-adsl]",
  path_outdata = tempfile(fileext = ".Rdata"),
  path_outtable = tempfile(fileext = ".rtf")
)
```

# Create Baseline Characteristic Table for Subgroup Analysis 
The baseline characteristic subgroup analysis aims to provide tables to summarize details of participants by subgroup.
The development of baseline characteristic subgroup analysis involves functions:

- `prepare_base_char_subgroup()`: prepare analysis raw datasets.
- `format_base_char_subgroup()`: prepare analysis (mock) outdata with proper format.
- `rtf_base_char_subgroup()`: transfer (mock) output dataset to RTF table.

## Analysis preparation

The `prepare_base_char_subgroup()` function is designed to be used for multiple purposes.
The input of the function is a `meta` object created by the metalite package.

```{r}
meta <- meta_sl_example()
```

The output of the function is an `outdata` object containing a list of analysis
raw datasets. Key arguments are:

- `subgroup_var`: a character value of subgroup variable name in observation data saved in meta$data_observation.
- `subgroup_header`: a character vector for column header hierarchy. The first element will be the first level header and the second element will be second level header.

```{r}
outdata <- prepare_base_char_subgroup(
  meta,
  population = "apat",
  parameter = "age",
  subgroup_var = "TRTA",
  subgroup_header = c("SEX","TRTA"),
  display_subgroup_total = TRUE
)
```

The output dataset contains commonly used statistics within each `subgroup_var`.

```{r}
outdata$out_all$`Placebo`
outdata$out_all$`High Dose`
outdata$out_all$`Low Dose`
```

The information about subgroup saved with `outdata$group` and `outdata$subgroup`.

```{r}
outdata$group
outdata$subgroup
```

`n_pop`: participants in population within each `subgroup_var`.

```{r}
outdata$out_all$`Placebo`$n
outdata$out_all$`High Dose`$n
outdata$out_all$`Low Dose`$n
```

## Format output

`format_base_char_subgroup` to prepare analysis dataset before generate RTF output

```{r}
tbl <- format_base_char_subgroup(outdata)
head(tbl$tbl)
```

## Output as RTF
`rtf_base_char_subgroup` to generate RTF output

```{r warning=FALSE}
outdata |>
  format_base_char_subgroup() |>
  rtf_base_char_subgroup(
    source = "Source:  [CDISCpilot: adam-adsl]",
    path_outdata = tempfile(fileext = ".Rdata"),
    path_outtable = tempfile(fileext = ".rtf")
  )
```

# Create Treatment Compliance Table
## Analysis preparation
```{r}
outdata <- meta |>
  prepare_trt_compliance()
outdata
```

## Output as RTF
```{r}
outdata |> 
  format_trt_compliance(display_col = c("n", "prop"), digits_prop = 3) |>
  rtf_trt_compliance(
  source = "Source: [CDISCpilot: adam-adsl]",
  path_outdata = tempfile(fileext = ".Rdata"),
  path_outtable = tempfile(fileext = ".rtf")
)
```

# Create Disposition Table
## Analysis preparation
```{r}
outdata <- meta |>
  prepare_disposition()
outdata
```

## Output as RTF
```{r, eval=FALSE}
outdata |> 
  format_disposition() |>
  rtf_disposition(
  source = "Source: [CDISCpilot: adam-adsl]",
  path_outdata = tempfile(fileext = ".Rdata"),
  path_outtable = tempfile(fileext = ".rtf")
)
```

